% This is the aspauthor.tex LaTeX file
% Copyright 2010, Astronomical Society of the Pacific Conference Series

\documentclass[11pt,twoside]{article}
\usepackage{asp2010}

\resetcounters

\bibliographystyle{asp2010}

\markboth{Bray}{Astropy}

\newcommand{\aspconf}{ASP Conf.\ Ser.}

\begin{document}

\title{Astropy: Building Blocks for Astronomy Software}

\author{Erik~Bray
\affil{Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA}}

\begin{abstract}
The Astropy Project is a community effort to develop an open source Python
package of common data structures and routines for use by other, more
specialized astronomy software in order to foster interoperability.
The project encompasses the "core" \texttt{astropy} Python package, "affiliated
packages" that strive to implement Astropy's coding standards and
interoperability with other affiliated packages, and a broader community aimed
at implementing Pythonic solutions to astronomy computing problems while
minimizing duplication of effort.  The project also provides a framework for
other projects that use Astropy to quickly get up and running by using the same
solutions Astropy has already developed for installation, packaging, testing,
documentation, and other nuts and bolts.
Here we present an overview of the key features of the core package (existing
and upcoming), current and planned affiliated packages, and how we manage a
large open source project with a diverse community of contributors.
\end{abstract}

\section{Core \texttt{astropy} Capabilities}

The \texttt{astropy} package for Python (referred to as the "core" package of
the Astropy Project) contains classes, functions, and utilities to address
common needs of other astronomy software as well as interactive data analysis.
It also includes a framework of sundry software development utilities such as a
test framework (built around py.test) documentation helpers using Sphinx, a
configuration system, and build and installation support built around distutils
and setuptools.

Here we list the sub-packages that make up the functionality of the core
\texttt{astropy} package as of version 0.2, as well as some features planned
for version 0.3 and beyond.

\subsection{Core Data Structures and Transformations}

\altsubsubsection*{\texttt{astropy.units}}

Handles defining and converting between physical units, and performing
arithmetic with physical quantities (numerical scalars and arrays with
associated units) using a powerful \texttt{Quantity} class.

\altsubsubsection*{\texttt{astropy.constants}}

Contains a number of physical constants useful in astronomy and physics.
Constants are represented as \texttt{Quantity} objects with additional metadata
describing their provenance and uncertainties.

\altsubsubsection*{\texttt{astropy.nddata}}

Provides an \texttt{NDData} class for managing n-dimensional array-based data
with attached metadata, errors, and masking.  Also provides new convolution
routines that treat NaN values properly.

\altsubsubsection*{\texttt{astropy.table}}

Provides a powerful \texttt{Table} class for storing and manipulating
heterogeneous tables of data in a way that is familiar to Numpy users, though
with more flexibility than the data structures built into Numpy including
support for metadata and masking.

\altsubsubsection*{\texttt{astropy.time}}
Provides functionality for manipulating times and dates.  Specific emphasis is
placed on supporting time scales (e.g.~UTC, TAI, UT1) and time representations
(e.g.~JD, MJD, ISO 8601) that are used in astronomy.  Wraps the ERFA library of
time and calendar routines, which is a fork of SOFA released under a less
restrictive license.

\altsubsubsection*{\texttt{astropy.coordinates}}

Provides classes and transformation functions for representing celestial
coordinates and for converting between standard systems in a uniform way.

\altsubsubsection*{\texttt{astropy.wcs}}

Contains utilities for managing WCS transformations in FITS files.  These
transformations map the pixel locations in an image to their real-world units.
This sub-package is an import of the already widely used PyWCS package.

\altsubsubsection*{\texttt{astropy.modeling}}

Provides a framework for representing models performing model evaluation and
fitting.  It supports 1D and 2D models and fitting with parameter constraints,
as well as an easy to use API for defining new models and fitting algorithms.

\subsection{File I/O}

One goal of Astropy is to minimize the effort of getting data into and out of
different file formats.  To that end it supports a registry of I/O "connectors"
that allow data from different formats to be transparently read into and
serialized from Astropy's general purpose \texttt{NDData} and \texttt{Table}
classes.  For example, a table can be transparently read from a FITS file,
modified as needed, and written out to an HDF5 file.  Support for lower-level
operations on specific file formats is maintained as well.

\altsubsubsection*{\texttt{astropy.io.fits}}

A port of the already widely used PyFITS package, supporting reading and
writing FITS files including support for many non-standard FITS conventions.

\altsubsubsection*{\texttt{astropy.io.ascii}}

A port of the popular asciitables package with support from reading from and
writing a wide range of plain text formats to and from Astropy \texttt{Table}
objects.

\altsubsubsection*{\texttt{astropy.io.votable}}

A port of the VOTable package for reading and writing the IVOA VOTable format
to and from Astropy \texttt{Table} objects, as well as validating them.

\altsubsubsection*{\texttt{astropy.io.misc}}

Includes optional support for tables stored in HD5 where h5py is available.

\subsection{Astronomy Computations and Utilities}

\altsubsubsection*{\texttt{astropy.cosmology}}

Contains classes for representing cosmologies, and utility functions for
calculating commonly used quantities that depend on a cosmological model.  This
includes distances, ages and lookback times corresponding to a measured
redshift or the transverse separation corresponding to a measured angular
separation.

\altsubsubsection*{\texttt{astropy.stats}}

This sub-package holds statistical functions and algorithms used in astronomy,
some of which are more specialized than one would find in
\texttt{scipy.stats}.

\altsubsubsection*{\texttt{astropy.vo}}

Handles simple access for Virtual Observatory (VO) services.  Currently, only
Simple Cone Search Version 1.03 as defined in IVOA Recommendation (February 22,
2008) is supported.

\section{Affiliated Packages}

The Astropy Project includes the concept of "affiliated packages".  An
affiliated package is an astronomy-related Python package that is not part of
the core \texttt{astropy} package, but has requested to be included under the
umbrella of the Astropy Project.  Some affiliated packages may be considered
for inclusion in the core package once they reach a reasonable level of
maturity and usefulness.  Others may be more specialized or have enough
external dependencies (GUI libraries, for example) that they make more sense
as separate packages.

The standards for inclusion in the Astropy Project are not strict, but it does
imply interoperability with other affiliated packages through use of the
common classes and data structures provided by the core package.  It also
implies adoption of similar coding standards as Astropy, as well as high
standards of documentation, testing, ease of installation, and open access.
To help achieve these standards we provide our framework in the form of a
package template that can be easily modified by upstart projects to suit
their needs, while providing some organizational homogeneity across affiliated
packages.

In the future we also plan to include simple tools for discovering and
installing new affiliated packages--we hope that becoming an Astropy
affiliated package will be a good way for new and existing Python astronomy
packages to gain exposure.

Some currently featured affiliated packages include Montage-wrapper, Ginga,
APLpy, astroML, and Astropysics.  A few additional affiliated packages that
are currently under development are photutils, astroquery, specutils, and
kcorrect.  The venerable pysynphot package is also being converted to an
Astropy affiliated package.

\section{Community and Collaboration}

The Astropy Project was launched as a response to an increasing proliferation
of Python packages (often with "astro" in their names) that attempt to build
core sets of routines ported from IDL and/or IRAF.  Most of these efforts were
supported by one or two individuals, or an individual institution, and did not
always involve much community support or feedback.  Some of us also became
concerned that too much effort was being spent reinventing the wheel, rather
than contributing to a common, open source code base.

A primary guiding principle behind Astropy has been that it is developed by and
for the astronomy community, and that it strives to include contributions from
users and other developers from around the world and from all sub-disciplines.
It also aims to encourage best practices from the software engineering world in
the hope of creating a clean, consistent, and stable API; thorough,
well-maintained, and easy to read documentation; and constant feedback through
careful issue tracking and continuous integration testing.  All of our
development practices and guidelines are documented as well so that other
projects--whether affiliated with Astropy or not--may adopt them without having
to develop their own standards from scratch.

In order to better enable collaboration we have adopted the GitHub open source
code hosting service which provides an issue tracker and a git source code
repository.  GitHub is designed around collaboration, and is set up to allow a
low barrier of entry to submitting new code and bug fixes to open source
projects.  Astropy has been "forked" over 100 times on GitHub, which is to say
over 100 people have their own copies of the Astropy repository in which they
can fix bugs and develop experimental new code to share back to the main
Astropy repository via "pull requests".  As of writing, over 50 contributors
have had their code pulled into the main Astropy repository--more than 20 with
over 10 commits and about 10 people with over 100 commits.  Over 200 people
have contributed with guidance, ideas, and issue reports via the GitHub issue
tracker and the astropy-dev mailing list.  Close to 1500 issues have been
opened on the issue tracker with more than 1000 of them resolved.

The development process ensures that any bug fixes or new features that are
implemented are accompanied by tests, and any relevant updates to the
documentation.  This keeps the documentation regularly up to date, even for the
development version of the software.  Free documentation building and hosting
by Read the Docs ensures that the latest documentation is always accessible.
All new pull requests are tested against multiple Python and Numpy versions
using the free continuous integration system Travis.  More comprehensive
testing against multiple OS platforms (Linux, MacOS X, and Windows) is made
possible by Jenkins with hosting from Shining Panda.  Continuous testing on
multiple platforms has made it possible for system integrators to include
Astropy in many other software package bundles such as Enthought Canopy,
Continuum Analytics' Anaconda, and OS packaging systems like Debian's apt.
Every single change made to the system is tested so that Astropy keeps working
on all platforms it supports, and is tracked so that users can easily find out
what changed between versions.

This development workflow has worked very well thus far, attracting
contributions from many new developers.  In fact the ease of contribution has
blurred the line between developers and users.  In fact, we have even
implemented a feature that allows anyone reading the documentation to submit
edits and improvements with just a few clicks in their web browser and no
prior knowledge of the git version control system.

It should be clarified that the Astropy Project is not purely democratic--it is
overseen by a coordinating committee consisting of Perry Greenfield (STScI,
which is committed significant developer resources to the project), Thomas
Robitaille (Max Planck Institute for Astronomy), and Erik Tollerud (Department
of Astronomy, Yale).  The coordinating committee drives the overall vision for
Astropy, and has the final vote on any disputes over design decisions.  In
practice, however, any contentions have been resolved through discussions on
GitHub and on the Astropy mailing list, resulting in designs that all
interested parties have agreed on.  We strive for continuous feedback and
rework, however, and hope to avoid a "designed by committee" feel.  So far, the
author would argue, we have been successful in this goal.

\bibliography{P95}

\end{document}
